### 2015-07-31 paper.tex
Add

\documentclass[11pt]{amsart}
\usepackage{setspace}
\usepackage{amssymb}
\usepackage{proof}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[thm]{Definition}
\newtheorem{example}[thm]{Example}
%\newtheorem{note}[thm]{Note}
\theoremstyle{remark}
\newtheorem{remark}[thm]{Remark}
\numberwithin{equation}{section}

\def\InputModeColorName{MidnightBlue}
\def\OutputModeColorName{Maroon}
\newcommand\InputMode[1]{{\color{\InputModeColorName}{#1}}}
\newcommand\OutputMode[1]{{\color{\OutputModeColorName}{#1}}}

\newcommand\HypJ[2]{#1\ (#2)}
\newcommand\GenJ[2]{\vert_{#1}\; #2}
\newcommand\JJ{\mathcal{J}}
\newcommand\IsVer[1]{\InputMode{#1}\ \mathit{ver}}
\newcommand\IsTrue[1]{\InputMode{#1}\ \mathit{true}}
\newcommand\IsProp[1]{\InputMode{#1}\ \mathit{prop}}
\newcommand\Member[2]{\InputMode{#1}\in \InputMode{#2}}
\newcommand\Seq[2]{\InputMode{#1}\vdash #2}
\newcommand\Nat{\mathbb{N}}

\begin{document}

\title{Remark on the hypothetical judgment}

\author{Jonathan Sterling}
\address{}

\onehalfspacing

\begin{abstract}

What is the proper explanation of intuitionistic hypothetical judgment, and
thence propositional implication? The answer to this question is not clearly
determinate from the writings of Brouwer and Heyting, since in their lifetimes
they propounded multiple (sometimes conflicting) explanations of the
hypothetical judgment. To my mind, the determination of an acceptable
explanation must be, then, an act of abduction and reconstruction, taking
into account its adequacy for the expression of the bar theorem and, more
generally, the development of an open-ended framework for transcendental
arguments in mathematics.

\end{abstract}

\maketitle

\section{Judgments and Propositions}

The distinction between the propositions and the judgments (assertions) is an
old one, but prior to Martin-L\"of, the significance of assertions was limited
to the affirmation of the truth of propositions. Following Martin-L\"of
\cite{siena.lectures}, forms of judgment other than $\IsTrue{P}$ are recognized,
including $\IsProp{P}$.

What is the difference between a judgment (assertion) on the one hand, and a
proposition on the other hand? A judgment is an act or an experience, whereas a
proposition is a mathematical object which may be experienced in different ways:
for instance, the assertion of the truth of a proposition (i.e.\ $\IsTrue{P}$)
consists in the fulfillment of the intention expressed by the proposition, and
the recognition of an object as a proposition (i.e.\ $\IsProp{P}$) is the act of
understanding this intention.

In addition to the categorical judgments above, higher-order forms of judgment
are also explained, including the hypothetical judgment and the general
judgment. Now, the primitive hypothetical judgment
$\HypJ{\JJ_2}{\JJ_1}$,\footnote{Hypothetical judgment is to be distinguished
from the sequent judgments $\Seq{\Gamma}{\cdots}$, which is not a higher-order
judgment at all.} was explained by Martin-L\"of in terms of his definition of a
hypothetical \emph{proof} or \emph{demonstration}, which is as follows:

\begin{quote}
  The notion of hypothetical proof [demonstration], in turn, which is a
primitive notion, is explained by saying that it is a proof [demonstration]
which, when supplemented by proofs [demonstrations] of the hypotheses, or
antecedents, becomes a proof [demonstration] of the thesis, or
consequent.\footnote{
I prefer the term \emph{demonstration} to the more ambiguous \emph{proof}, since
the former is clearly an act, whereas the latter may be read as either an act or
as a mathematical object.
}
\cite{siena.lectures}
\end{quote}

In 1956, Heyting propounded his version of what has come to be known as the
\emph{Brouwer-Heyting-Kolmogorov} interpretation of intuitionistic logic, by
explaining the assertion conditions of the propositions. Note that where Heyting
says ``assert a proposition'', in light of Martin-L\"of's clarification, we must
understand ``assert \emph{the truth of} a proposition''. Heyting's explanation
of the assertion conditions for the truth of implication were as follows:
\begin{quote}
  The implication $p\to q$ may be asserted if and only if we possess a
  construction $r$, which, joined to any construction proving $p$ (supposing the
  latter be effected), would automatically effect a construction of $q$. \cite{HeytingA:int}
\end{quote}

Now, Martin-L\"of would probably consider the parenthetical ``supposing the
latter be effected'' to be superfluous, since to assume a judgment is the same
as to assume that you know it \cite{siena.lectures}.
%%% note that this is not obvious to people who are not PML
So, Heyting's definition might be rewritten today as:

\begin{quote}
$\IsTrue{P\to Q}$ may be asserted if and only if we
possess a construction $r$, which, joined to any demonstration of $\IsTrue{P}$,
would automatically effect a demonstration of $\IsTrue{Q}$.
\end{quote}

Martin-L\"of explained the truth of an implication by appealing to the
hypothetical judgment, so we should be able to factor Heyting's explanation
through it in a similar way:

\begin{quote}
  $\IsTrue{P\to Q}$ may be asserted if and only if we may assert $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
\end{quote}

In fact, if we make this transformation, we shall have arrived at something very
similar to Martin-L\"of's definition of propositional implication. This
inference is, at least, valid with respect to Martin-L\"of's definition, but it
is merely an extensional specification for the meaning of the judgment: it
expresses the material equivalence of the assertions $\IsTrue{P\to Q}$ and
$\HypJ{\IsTrue{Q}}{\IsTrue{P}}$, but it does not contain an actual
\emph{explanation} of $\IsTrue{P\to Q}$, which would need to be in the form ``To
know $\IsTrue{P\to Q}$ is to know...''.

We cannot actually rewrite our definition directly in that style because it is
impredicative as a definition (as opposed to as a mathematical statement);
following Dummett \cite{Dummett:Elements} and Martin-L\"of, we must start from a
distinction between the direct experience of truth (which we shall call
\emph{verification}) and the indirect experience (which we shall call merely
\emph{truth}). Then, the intention of a proposition is its verification, and the
truth of a proposition is the experience of a means of fulfillment for that
intention:

\begin{quote}
  To know $\IsVer{P\to Q}$ is to know $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
\end{quote}

Then, to know $\IsTrue{P\to Q}$ is to have a means of verifying $P\to Q$, that
is, to have a plan to experience $\IsVer{P\to Q}$.

\section{The Proof Interpretation}

When we rewrote Heyting's explanation of the assertion conditions for the truth
of an implication to appeal to hypothetical judgment, we implicitly assumed that
the instantiation of Martin-L\"of's hypothetical judgment would preserve the
meaning of the original statement.

This, however, may be too much to ask, since in modern proof-theoretic accounts
of meaning, the hypothetical judgment as explained by Martin-L\"of must be
understood in a very strong sense, where its proof shall be an object with a
``hole'' in it, which could be plugged with a proof for the antecedent to yield
a proof for the consequent: that is, it is not enough that one should have a way
of transforming the proof of the antecedent into proof for the consequent, but
one must have a uniform way to do so.

Anticipating the ``proof interpretation'' of intuitionistic logic, Brouwer also
had come to a conclusion similar to this, if not quite equivalent. For Brouwer,
a hypothetical assertion $\HypJ{\JJ_2}{\JJ_1}$ was essentially an assertion of
$\JJ_2$ which proceeds by embedding an actual construction of $\JJ_1$ into a
matrix for a construction of $\JJ_2$. Now, this is not quite the same since it
requires that $\JJ_1$ be proved, so it corresponds more closely with the notion
of \emph{cut} than it does with hypothetical judgment---but he does seem to
agree as far as the uniformity of the partial proof of the consequent is
concerned.

In case it is not entirely evident, let us reason through what would happen to a
\emph{proof theory} if proofs of hypothetical judgments could be non-uniform.
For one, it would cease to be a proof theory, since two crucial properties would
fail:

\begin{enumerate}
  \item Proofs are finitary objects.
  \item It is effectively determinable in finite time whether an object is
    a proof of a judgment.
\end{enumerate}

If a proof of $\HypJ{\JJ_2}{\JJ_1}$ were construed as a means of converting
proofs of $\JJ_1$ into proofs of $\JJ_2$ (as opposed to the proof-with-a-hole
interpretation), proofs would certainly cease to be finitary objects: for
instance, a proof of $\GenJ{n}{\HypJ{\JJ(n)}{\Member{n}{\Nat}}}$ would be
infinitely large, in that it would contain one branch for each natural number.
Then, the failure of the second property (decidability of wellformedness) is
immediate.

Proof-theoretic meaning, however, is hopeless anyway as an interpretation of
intuitionistic logic if one has any intention to justify the creating subject
\cite{mva:personal}, or even the bar theorem; indeed, the proof-theoretic
reading of hypothetical judgment specifically rules out the kind of non-uniform
evidence that is essential in the demonstration of the bar theorem.

\section{The Proper Interpretation of Hypothetical Judgment}

If we return to Heyting's original definition of the assertion $\IsTrue{P\to
Q}$, it is clear that he would have to accept any mathematical means of
transforming the construction of $\IsTrue{P}$ into a construction of
$\IsTrue{Q}$, since the construction $r$ in his definition is not constrained by
any uniformity condition: it is merely any effective operation which, when
adjoined with a construction of the premise, effects a construction of the
conclusion; and moreover, $r$ may proceed by virtue of any of the transcendental
observations which are possible by virtue of the truth of $P$ having been
experienced by the subject.

And what of Martin-L\"of? In his type theory, only the introduction rules for
the types are given, and the elimination rules are ``theorems'' (or admissible
rules) which are evident only in the non-uniform sense; that is, the
justifications of the elimination rules proceed by introspection on the possible
ways in which their premises were experienced. Therefore, if Martin-L\"of's
uniform hypothetical judgment is to be accepted, it is necessary that the
statement of the elimination rules be effected using something that permits
a non-uniform demonstration.

Sundholm and Van Atten \cite{sundholm-mva}, for instance, distinguish between a
hypothetical proof of $\HypJ{\JJ_2}{\JJ_1}$ and an inference
$\infer{\JJ_2}{\JJ_1}$ which expresses the closure of mathematics under a rule,
and say that it is the latter which establishes the bar theorem, and not
the former.

Martin-L\"of on the other hand says specifically that an inference is to be read
as a proof of a hypothetical judgment, and so no real progress is made:
\begin{quote}
  The difference between an inference and a logical consequence, or hypothetical
  judgement, is that an inference is a proof of a logical consequence. Thus an
  inference is the same as a hypothetical proof. \cite{siena.lectures}
\end{quote}

In light of Martin-L\"of's explanations of the elimination rules which follow
this statement, we cannot accept his claim that inference is the same as
hypothetical proof (as he has defined it), because under that definition even
the following is not a valid inference:
\[
  \infer{
    \IsTrue{P}
  }{
    \IsTrue{P\land Q}
  }
\]

Why? This purports to be a proof of the judgment
$\HypJ{\IsTrue{P}}{\IsTrue{P\land Q}}$, which is not evident under the uniform
explanation of hypothetical judgment. It could be made evident if the definition
of a proposition consisted in the declaration of both its introduction rules and
its elimination rules, but in fact, only the introduction rules are given, and
the elimination rules are simply codifications of common patterns of inference.

Indeed, Martin-L\"of justifies the elimination rules using the non-uniform
notion of consequence. This rule, for instance, is justified as follows:

\begin{proof}
If you know $\IsTrue{P\land Q}$, then you must have a means of verifying $P\land
Q$, whence you must know both $\IsTrue{P}$ and $\IsTrue{Q}$; the conclusion is
now immediate.
\end{proof}

If we are to take the explanation of the elimination rules seriously, then, we
must read Martin-L\"of as having already a kind of hypothetical judgment whose
evidence consists in any effective means at all of transforming the evidence for
the premise into the evidence for the conclusion (we might call this ``material
consequence'' as opposed to ``logical consequence''), so the use of the uniform
hypothetical judgment elsewhere has not somehow removed the need for this kind
of inference.

The uniform explanation of hypothetical judgment also deprives Martin-L\"of's
framework of a propositional connective as strong as proper intuitionistic
implication. In spite of Martin-L\"of's statement in the Siena lectures, the
interpretation of hypothetical judgment as material consequence is also crucial
for the semantics of his type theory, as noted by Dybjer \cite{dybjer:testing}.

%%% TODO:
%%% - discuss propositional witnesses
%%% - realizability as unary logical relations
%%% - (nuprl) type theory as binary logical relations
%%% - what do the propositions give us over the judgments? (witnesses, extensional identity)
%%% - harper's presentation of admissibility & derivability in pfpl
%%% - noam zeilberger's use of non-uniform hypothetical judgment

\bibliographystyle{plain}
\bibliography{refs}

\end{document}

### 2015-08-01 paper.tex
wording

@@
 framework of a propositional connective as strong as proper intuitionistic
-implication. In spite of Martin-L\"of's statement in the Siena lectures, the
+implication. Contra Martin-L\"of's statement in the Siena lectures, the
 interpretation of hypothetical judgment as material consequence is also crucial

### 2015-08-01 paper.tex
begin discussing propositional witnesses

@@
 If we are to take the explanation of the elimination rules seriously, then, we
-must read Martin-L\"of as having already a kind of hypothetical judgment whose
-evidence consists in any effective means at all of transforming the evidence for
-the premise into the evidence for the conclusion (we might call this ``material
-consequence'' as opposed to ``logical consequence''), so the use of the uniform
-hypothetical judgment elsewhere has not somehow removed the need for this kind
-of inference.
+must read Martin-L\"of as having already at his disposal a kind of hypothetical
+judgment whose evidence consists in any effective means at all of transforming
+the evidence for the premise into the evidence for the conclusion (we might call
+this ``material consequence'' as opposed to ``logical consequence''), so the use
+of the uniform hypothetical judgment elsewhere has not somehow removed the need
+for this kind of inference.
 
@@
 
+\section{Realizability and Type Theory}
+
+Now that we have settled upon an explanation for hypothetical judgment, let us
+return to the notion of propositions and implication, and their respective
+notions of ``construction''. Like ``proof'', the term ``construction'' is also
+ambiguous in that it may refer to an act of constructing, and it may also denote
+a concrete mathematical object.
+
+A construction for a judgment is simply the act of coming to know it---if it is
+to be thought of as an object, it is at least a tensed, ephemeral one; this is
+what Martin-L\"of calls a \emph{demonstration}. On the other hand, a
+construction for a proposition is a mathematical object, not an experience: it
+is the object that the subject constructs during the verification of a
+proposition. This latter sort of construction is called a \emph{witness}, or,
+following realizability, a \emph{realizer}.
+
 %%% TODO:
-%%% - discuss propositional witnesses
 %%% - realizability as unary logical relations

### 2015-08-01 paper.tex
explain unary logical relations

@@
 \usepackage{setspace}
-\usepackage{amssymb}
+\usepackage{amssymb,stmaryrd}
 \usepackage{proof}
@@
 \newcommand\IsProp[1]{\InputMode{#1}\ \mathit{prop}}
+\newcommand\IsSet[1]{\InputMode{#1}\ \mathit{set}}
 \newcommand\Member[2]{\InputMode{#1}\in \InputMode{#2}}
 \newcommand\Seq[2]{\InputMode{#1}\vdash #2}
+\newcommand\Eval[2]{\InputMode{#1}\Rightarrow \OutputMode{#2}}
 \newcommand\Nat{\mathbb{N}}
+
+\newcommand\SEM[1]{\llbracket#1 \rrbracket}
+\newcommand\VAL[1]{\mathcal{V}\SEM{#1}}
+\newcommand\EXP[1]{\mathcal{E}\SEM{#1}}
+
+\newcommand\True{\top}
+\newcommand\False{\bot}
+\newcommand\Imp[2]{#1\supset #2}
+\newcommand\Conj[2]{#1\land #2}
+\newcommand\It{\star}
+\newcommand\Lam[2]{(\lambda #1) #2}
+\newcommand\Pair[2]{\langle #1, #2 \rangle}
+\newcommand\Fst{\mathsf{fst}}
+\newcommand\Ap{\mathsf{ap}}
+\newcommand\IsDefined[1]{\InputMode{#1}\ \mathit{defined}}
 
@@
 \begin{quote}
-  The implication $p\to q$ may be asserted if and only if we possess a
+  The implication $\Imp{p}{q}$ may be asserted if and only if we possess a
   construction $r$, which, joined to any construction proving $p$ (supposing the
@@
 \begin{quote}
-$\IsTrue{P\to Q}$ may be asserted if and only if we
+$\IsTrue{\Imp{P}{Q}}$ may be asserted if and only if we
 possess a construction $r$, which, joined to any demonstration of $\IsTrue{P}$,
@@
 \begin{quote}
-  $\IsTrue{P\to Q}$ may be asserted if and only if we may assert $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
+  $\IsTrue{\Imp{P}{Q}}$ may be asserted if and only if we may assert $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
 \end{quote}
@@
 is merely an extensional specification for the meaning of the judgment: it
-expresses the material equivalence of the assertions $\IsTrue{P\to Q}$ and
+expresses the material equivalence of the assertions $\IsTrue{Imp{P}{Q}}$ and
 $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$, but it does not contain an actual
-\emph{explanation} of $\IsTrue{P\to Q}$, which would need to be in the form ``To
-know $\IsTrue{P\to Q}$ is to know...''.
+\emph{explanation} of $\IsTrue{Imp{P}{Q}}$, which would need to be in the form ``To
+know $\IsTrue{Imp{P}{Q}}$ is to know...''.
 
@@
 \begin{quote}
-  To know $\IsVer{P\to Q}$ is to know $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
+  To know $\IsVer{Imp{P}{Q}}$ is to know $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
 \end{quote}
 
-Then, to know $\IsTrue{P\to Q}$ is to have a means of verifying $P\to Q$, that
-is, to have a plan to experience $\IsVer{P\to Q}$.
+Then, to know $\IsTrue{Imp{P}{Q}}$ is to have a means of verifying $Imp{P}{Q}$, that
+is, to have a plan to experience $\IsVer{Imp{P}{Q}}$.
 
@@
 
-If we return to Heyting's original definition of the assertion $\IsTrue{P\to
-Q}$, it is clear that he would have to accept any mathematical means of
-transforming the construction of $\IsTrue{P}$ into a construction of
+If we return to Heyting's original definition of the assertion
+$\IsTrue{\Imp{P}{Q}}$, it is clear that he would have to accept any mathematical
+means of transforming the construction of $\IsTrue{P}$ into a construction of
 $\IsTrue{Q}$, since the construction $r$ in his definition is not constrained by
@@
 
+\subsection{Realizability Models as Unary Logical Relations}
+
+In fact, by replacing the abstract/synthetic explanations of the propositions in
+terms of their verification acts, we can recast them as giving rise to a theory
+of sets or types by explaining them in terms of their verification objects (i.e.
+their canonical witnesses).
+
+A realizability model in this simple sense amounts to interpreting the
+propositions into unary logical relations. Defining a proposition, then, amounts
+to defining the unary relation $\VAL{P}$ (which is the set of verification
+objects of the proposition $P$); then, a separate logical relation $\EXP{P}$ is
+defined uniformly over all propositions $P$ by appealing to the computation
+$\Eval{M}{N}$ of witnesses to canonical form.
+\begin{align*}
+  \VAL{\True} &\equiv \{\It\}\\
+  \VAL{\False} &\equiv \{\}\\
+  \VAL{\Imp{P}{Q}} &\equiv 
+    \{ \Lam{x}{E}
+    \mid \GenJ{x}{\HypJ{\Member{E}{\EXP{Q}}}{\Member{x}{\EXP{P}}}}
+    \}\\
+  \VAL{\Conj{P}{Q}} &\equiv
+    \{ \Pair{M}{N}
+    \mid \Member{M}{\EXP{P}}, \Member{N}{\EXP{Q}}
+    \}\\
+  \EXP{P} &\equiv
+    \{ M
+    \mid \Eval{M}{M'}, \Member{M'}{\VAL{P}}
+    \}
+\end{align*}
+
+Then, the judgments of the logical theory are explained by appealing to the
+logical relations:
+
+\begin{gather*}
+  \infer={
+    \IsProp{P}
+  }{
+    \IsDefined{\VAL{P}}
+  } \qquad
+  \infer{
+    \IsVer{P}
+  }{
+    \Member{M}{\VAL{P}}
+  }\qquad
+  \infer{
+    \IsTrue{P}
+  }{
+    \Member{M}{\EXP{P}}
+  }
+\end{gather*}
+
+The judgments of a basic theory of sets may also be explained using the logical
+relations:
+\begin{gather*}
+  \infer={
+    \IsSet{A}
+  }{
+    \IsDefined{\VAL{A}}
+  } \qquad
+  \infer={
+    \Member{M}{A}
+  }{
+    \Member{M}{\EXP{A}}
+  }
+\end{gather*}
+
+The material interpretation of the hypothetical judgment is crucial in the
+realizability model; this is because only the verification objects (i.e.
+canonical witnesses) are given. All the non-canonical witnesses are explained
+via computation, and the use of logical consequence instead of material
+consequence in the explanation of $\VAL{\Imp{P}{Q}}$ would have been disastrous.
+
+For instance, it should be the case that $\Lam{x}{\Pair{\It}{\It}}$ is a witness
+of $\Imp{\False}{\True}$. To see if this is the case in the
+model, let us translate this into a concrete statement:
+\begin{gather}
+  \Member{\Lam{x}{\Pair{\It}{\It}}}{\EXP{\Imp{\False}{\True}}}\\
+  \Eval{\Lam{x}{\Pair{\It}{\It}}}{\Lam{x}{\Pair{\It}{\It}}}, \quad
+    \Member{\Lam{x}{\Pair{\It}{\It}}}{\VAL{\Imp{\False}{\True}}}\\
+  \GenJ{x}{
+    \HypJ{\Member{\Pair{\It}{\It}}{\EXP{\True}}}{\Member{x}{\EXP{\False}}}
+  }\\
+  \GenJ{x}{
+    \HypJ{\Member{\Pair{\It}{\It}}{\EXP{\True}}}{\Eval{x}{M}, \Member{M}{\VAL{\False}}}
+  }
+\end{gather}
+
+And at this time, we may discharge the entire hypothetical judgment, since we
+know that the unary relation $\VAL{\False}$ is empty, and so there can be no
+such $M$.
+
+If we did not have the material consequence at our disposal, then this statement
+would not have been valid, unless we were to eschew the verificationist meaning
+explanation and also add ``use'' rules (i.e. direct elimination) for each
+proposition in addition to the verification rules.
+
 %%% TODO:
-%%% - realizability as unary logical relations
 %%% - (nuprl) type theory as binary logical relations

### 2015-08-01 paper.tex
typo
Add section on binary logical relations

@@
 \documentclass[11pt]{amsart}
-\usepackage{setspace}
+\usepackage{setspace,url,hyperref}
 \usepackage{amssymb,stmaryrd}
@@
 \newcommand\IsSet[1]{\InputMode{#1}\ \mathit{set}}
+\newcommand\EqSet[2]{\InputMode{#1}=\InputMode{#2}\ \mathit{set}}
 \newcommand\Member[2]{\InputMode{#1}\in \InputMode{#2}}
+\newcommand\EqMember[3]{\InputMode{#1}=\InputMode{#2}\in \InputMode{#3}}
 \newcommand\Seq[2]{\InputMode{#1}\vdash #2}
+\newcommand\SSeq[2]{\InputMode{#1}\vDash #2}
 \newcommand\Eval[2]{\InputMode{#1}\Rightarrow \OutputMode{#2}}
-\newcommand\Nat{\mathbb{N}}
+\newcommand\IsDefined[1]{\InputMode{#1}\ \mathit{defined}}
 
@@
 
+\newcommand\Nat{\mathbb{N}}
 \newcommand\True{\top}
@@
 \newcommand\Ap{\mathsf{ap}}
-\newcommand\IsDefined[1]{\InputMode{#1}\ \mathit{defined}}
 
@@
 $\Eval{M}{N}$ of witnesses to canonical form.
-\begin{align*}
-  \VAL{\True} &\equiv \{\It\}\\
-  \VAL{\False} &\equiv \{\}\\
-  \VAL{\Imp{P}{Q}} &\equiv 
-    \{ \Lam{x}{E}
-    \mid \GenJ{x}{\HypJ{\Member{E}{\EXP{Q}}}{\Member{x}{\EXP{P}}}}
-    \}\\
-  \VAL{\Conj{P}{Q}} &\equiv
-    \{ \Pair{M}{N}
-    \mid \Member{M}{\EXP{P}}, \Member{N}{\EXP{Q}}
-    \}\\
-  \EXP{P} &\equiv
-    \{ M
-    \mid \Eval{M}{M'}, \Member{M'}{\VAL{P}}
-    \}
-\end{align*}
 
-Then, the judgments of the logical theory are explained by appealing to the
-logical relations:
-
-\begin{gather*}
-  \infer={
-    \IsProp{P}
-  }{
-    \IsDefined{\VAL{P}}
-  } \qquad
-  \infer{
-    \IsVer{P}
-  }{
-    \Member{M}{\VAL{P}}
-  }\qquad
-  \infer{
-    \IsTrue{P}
-  }{
-    \Member{M}{\EXP{P}}
-  }
-\end{gather*}
-
-The judgments of a basic theory of sets may also be explained using the logical
-relations:
 \begin{gather*}
@@
   }{
-    \IsDefined{\VAL{A}}
+    \Eval{A}{A'} &
+    \IsDefined{\VAL{A'}}
   } \qquad
@@
   }{
-    \Member{M}{\EXP{A}}
+    \Eval{A}{A'} &
+    \EXP{A'}(\InputMode{M})
   }
 \end{gather*}
+\begin{align*}
+  \VAL{\True} &\equiv \{\InputMode{\It}\}\\
+  \VAL{\False} &\equiv \{\}\\
+  \VAL{\Imp{P}{Q}} &\equiv 
+    \{ \InputMode{\Lam{x}{E}}
+    \mid \GenJ{x}{\HypJ{\Member{E}{Q}}{\Member{x}{P}}}
+    \}\\
+  \VAL{\Conj{P}{Q}} &\equiv
+    \{ \InputMode{\Pair{M}{N}}
+    \mid \Member{M}{P}, \Member{N}{Q}
+    \}\\
+  \EXP{P} &\equiv
+    \{ \InputMode{M}
+    \mid \Eval{M}{M'}, \VAL{P}(\InputMode{M'})
+    \}
+\end{align*}
 
@@
 \begin{gather}
-  \Member{\Lam{x}{\Pair{\It}{\It}}}{\EXP{\Imp{\False}{\True}}}\\
+  \Member{\Lam{x}{\Pair{\It}{\It}}}{\Imp{\False}{\True}}\\
   \Eval{\Lam{x}{\Pair{\It}{\It}}}{\Lam{x}{\Pair{\It}{\It}}}, \quad
-    \Member{\Lam{x}{\Pair{\It}{\It}}}{\VAL{\Imp{\False}{\True}}}\\
+    \VAL{\Imp{\False}{\True}}(\InputMode{\Lam{x}{\Pair{\It}{\It}}})\\
   \GenJ{x}{
-    \HypJ{\Member{\Pair{\It}{\It}}{\EXP{\True}}}{\Member{x}{\EXP{\False}}}
+    \HypJ{\Member{\Pair{\It}{\It}}{\True}}{\Member{x}{\False}}
   }\\
   \GenJ{x}{
-    \HypJ{\Member{\Pair{\It}{\It}}{\EXP{\True}}}{\Eval{x}{M}, \Member{M}{\VAL{\False}}}
+    \HypJ{\Member{\Pair{\It}{\It}}{\True}}{\Eval{x}{M}, \VAL{\False}(\InputMode{M})}
   }
@@
 would not have been valid, unless we were to eschew the verificationist meaning
-explanation and also add ``use'' rules (i.e. direct elimination) for each
+explanation and also add ``use'' rules (i.e. direct eliminations) for each
 proposition in addition to the verification rules.
 
+\subsection{Type Theories as Binary Logical Relations}
+
+The unary logical relations exactly express the content of a realizability
+model, but they still do not yield a theory of sets which is sufficient for
+reasoning about mathematical objects, which have extensional identity. In order
+to consider the equality of of witnesses, the unary logical relations are
+replaced with binary ones (i.e. partial equivalence relations), as follows:
+
+\begin{gather*}
+  \infer={
+    \EqSet{A}{B}
+  }{
+    \Eval{A}{A'} &
+    \Eval{B}{B'} &
+    \InputMode{\VAL{A'}}\equiv\InputMode{\VAL{B'}}
+  }\qquad
+  \infer={
+    \IsSet{A}
+  }{
+   \EqSet{A}{A}
+  }\\
+  \infer={
+    \EqMember{M}{N}{A}
+  }{
+    \Eval{A}{A'} &
+    \Eval{M}{M'} &
+    \Eval{N}{N'} &
+    \VAL{A'}(\InputMode{M}, \InputMode{N})
+  }\qquad
+  \infer={
+    \Member{M}{A}
+  }{
+    \EqMember{M}{M}{A}
+  }
+\end{gather*}
+
+\begin{align*}
+  \VAL{\True} &\equiv \{(\InputMode{\It}, \InputMode{\It})\}\\
+  \VAL{\False} &\equiv \{\}\\
+  \VAL{\Imp{P}{Q}} &\equiv 
+    \{ (\InputMode{\Lam{x}{E}}, \InputMode{\Lam{x}{E'}})
+    \mid \GenJ{y,z}{
+      \HypJ{\EqMember{[y/x]E}{[z/x]E'}{Q}}{\EqMember{x}{y}{P}}
+    }
+    \}\\
+  \VAL{\Conj{P}{Q}} &\equiv
+    \{ (\InputMode{\Pair{M}{N}}, \InputMode{\Pair{M'}{N'}})
+    \mid \EqMember{M}{M'}{P}, \EqMember{N}{N'}{Q}
+    \}\\
+  \EXP{P} &\equiv
+    \{ (\InputMode{M}, \InputMode{N})
+    \mid \Eval{M}{M'}, \Eval{N}{N'}, \VAL{P}(\InputMode{M'}, \InputMode{N'})
+    \}
+\end{align*}
+
+With the extension to the binary logical relation, we now properly treat the
+equivalence of propositions (sets) and of witnesses, and we also have a
+definitive answer to the question, ``What is the purpose of adding a language of
+types and witnesses to the existing system of judgments and their
+demonstrations?''
+
+The judgments and demonstration acts are the activity of the subject in
+performing and experiencing mathematics. On the other hand, the theory of sets
+that we have defined is the correct level at which to \emph{do} mathematics,
+where objects are concrete and have an extensional identity. By guaranteeing
+pervasive functionality, we have embedded in the rich world of intuitionistic
+mathematics a haven safe from the paradoxes that arise from failures of
+extensionality, such as Diaconescu's theorem \cite{sterling:diaconescu}.
+
 %%% TODO:
-%%% - (nuprl) type theory as binary logical relations
-%%% - what do the propositions give us over the judgments? (witnesses, extensional identity)
 %%% - harper's presentation of admissibility & derivability in pfpl
@@
 
+\newpage
 \bibliographystyle{plain}

### 2015-08-01 paper.tex
add discussion of harper & zeilberger stuff

@@
 
-%%% TODO:
-%%% - harper's presentation of admissibility & derivability in pfpl
-%%% - noam zeilberger's use of non-uniform hypothetical judgment
+\section{Related Discussion}
+
+In the logical framework which forms the basis of \emph{Practical Foundations
+for Programming Languages} \cite{PFPL}, Robert Harper treats both logical
+consequence $\JJ_1\vdash\JJ_2$ and material consequence $\JJ_1\vDash\JJ_2$,
+which express derivability and admissiblity respectively.
+
+The material consequence (and its open-ended interpretation as a mapping from
+demonstrations of the antecedent to demonstrations of the consequent) formed the
+backbone of Zeilberger, Harper and Licata's work on focused calculi, which
+provide a convincing abstract notation for the traces of verification and use
+acts in a logic which mixes the verificationist and pragmatist meaning
+explanations \cite{zeilberger:thesis, licata-zeilberger-harper:focusing,
+zeilberger:2008}.
 

### 2015-08-01 paper.tex
add acknowledgement
typo

@@
 is merely an extensional specification for the meaning of the judgment: it
-expresses the material equivalence of the assertions $\IsTrue{Imp{P}{Q}}$ and
+expresses the material equivalence of the assertions $\IsTrue{\Imp{P}{Q}}$ and
 $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$, but it does not contain an actual
-\emph{explanation} of $\IsTrue{Imp{P}{Q}}$, which would need to be in the form ``To
-know $\IsTrue{Imp{P}{Q}}$ is to know...''.
+\emph{explanation} of $\IsTrue{\Imp{P}{Q}}$, which would need to be in the form ``To
+know $\IsTrue{\Imp{P}{Q}}$ is to know...''.
 
@@
 \begin{quote}
-  To know $\IsVer{Imp{P}{Q}}$ is to know $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
+  To know $\IsVer{\Imp{P}{Q}}$ is to know $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
 \end{quote}
 
-Then, to know $\IsTrue{Imp{P}{Q}}$ is to have a means of verifying $Imp{P}{Q}$, that
-is, to have a plan to experience $\IsVer{Imp{P}{Q}}$.
+Then, to know $\IsTrue{\Imp{P}{Q}}$ is to have a means of verifying $\Imp{P}{Q}$, that
+is, to have a plan to experience $\IsVer{\Imp{P}{Q}}$.
 
@@
 
+\subsubsection*{Acknowledgements}
+
+Thanks to Mark van Atten, Robert Harper, Clarissa Littler, and Danny Gratzer for
+helpful conversations about hypothetical judgment.
+
 \newpage

### 2015-08-01 paper.tex
typo

@@
 Now that we have settled upon an explanation for hypothetical judgment, let us
-return to the notion of propositions and implication, and their respective
-notions of ``construction''. Like ``proof'', the term ``construction'' is also
-ambiguous in that it may refer to an act of constructing, and it may also denote
-a concrete mathematical object.
+return to the notions of judgment and proposition, and their respective concepts
+of ``construction''. Like ``proof'', the term ``construction'' is also ambiguous
+in that it may refer to an act of constructing, and it may also denote a
+concrete mathematical object.
 

### 2015-08-01 paper.tex
typo

@@
 $\HypJ{\JJ_2}{\JJ_1}$,\footnote{Hypothetical judgment is to be distinguished
-from the sequent judgments $\Seq{\Gamma}{\cdots}$, which is not a higher-order
-judgment at all.} was explained by Martin-L\"of in terms of his definition of a
-hypothetical \emph{proof} or \emph{demonstration}, which is as follows:
+from the sequent judgments $\Seq{\Gamma}{\cdots}$, which are not even
+higher-order judgments at all.} was explained by Martin-L\"of in terms of his
+definition of a hypothetical \emph{proof} or \emph{demonstration}, which is as
+follows:
 

### 2015-08-01 paper.tex
more references
...

@@
 Thanks to Mark van Atten, Robert Harper, Clarissa Littler, and Danny Gratzer for
-helpful conversations about hypothetical judgment.
+helpful conversations about hypothetical judgment, elimination rules and the bar
+theorem.
 
 \newpage
+
+\nocite{*}
 \bibliographystyle{plain}

### 2015-08-01 paper.tex
fix typos; clarify ontological status of verification objects

@@
 \begin{quote}
-  The implication $\Imp{p}{q}$ may be asserted if and only if we possess a
-  construction $r$, which, joined to any construction proving $p$ (supposing the
-  latter be effected), would automatically effect a construction of $q$. \cite{HeytingA:int}
+  The implication $\Imp{\mathfrak{p}}{\mathfrak{q}}$ may be asserted if and only
+if we possess a construction $\mathfrak{r}$, which, joined to any construction
+proving $\mathfrak{p}$ (supposing the latter be effected), would automatically
+effect a construction of $\mathfrak{q}$. \cite{HeytingA:int}
 \end{quote}
@@
 a proof for the consequent: that is, it is not enough that one should have a way
-of transforming the proof of the antecedent into proof for the consequent, but
+of transforming the proof of the antecedent into a proof of the consequent, but
 one must have a uniform way to do so.
@@
 means of transforming the construction of $\IsTrue{P}$ into a construction of
-$\IsTrue{Q}$, since the construction $r$ in his definition is not constrained by
-any uniformity condition: it is merely any effective operation which, when
-adjoined with a construction of the premise, effects a construction of the
-conclusion; and moreover, $r$ may proceed by virtue of any of the transcendental
-observations which are possible by virtue of the truth of $P$ having been
-experienced by the subject.
+$\IsTrue{Q}$, since the construction $\mathfrak{r}$ in his definition is not
+constrained by any uniformity condition: it is merely any effective operation
+which, when adjoined with a construction of the premise, effects a construction
+of the conclusion; and moreover, $\mathfrak{r}$ may proceed by virtue of any of
+the transcendental observations which are possible by virtue of the truth of $P$
+having been experienced by the subject.
 
@@
 justifications of the elimination rules proceed by introspection on the possible
-ways in which their premises were experienced. Therefore, if Martin-L\"of's
-uniform hypothetical judgment is to be accepted, it is necessary that the
-statement of the elimination rules be effected using something that permits
-a non-uniform demonstration.
+ways in which their premises could have been experienced. Therefore, if
+Martin-L\"of's uniform hypothetical judgment is to be accepted, it is necessary
+that the statement of the elimination rules be effected using something that
+permits a non-uniform demonstration.
 
@@
 its elimination rules, but in fact, only the introduction rules are given, and
-the elimination rules are simply codifications of common patterns of inference.
+the elimination rules are simply codifications of common patterns of reasoning
+from premise to conclusion.
 
 Indeed, Martin-L\"of justifies the elimination rules using the non-uniform
-notion of consequence. This rule, for instance, is justified as follows:
+(material, rather than logical) consequence. This rule, for instance, is
+justified as follows:
 
@@
 judgment whose evidence consists in any effective means at all of transforming
-the evidence for the premise into the evidence for the conclusion (we might call
-this ``material consequence'' as opposed to ``logical consequence''), so the use
-of the uniform hypothetical judgment elsewhere has not somehow removed the need
-for this kind of inference.
+the demonstration of the premise into a demonstration of the conclusion, so the
+use of the uniform hypothetical judgment elsewhere has not somehow saved us the
+need to explain this kind of inference.
 
@@
 
-In fact, by replacing the abstract/synthetic explanations of the propositions in
-terms of their verification acts, we can recast them as giving rise to a theory
-of sets or types by explaining them in terms of their verification objects (i.e.
-their canonical witnesses).
+In fact, we can replace the abstract/synthetic explanations of the propositions
+in terms of their verification acts with new explanations in terms of
+verification objects (i.e. their canonical witnesses); then, the verification
+act consists in constructing a verification object.
+
+A proposition is verified just in case there exists a verification object, but
+it is important to understand that this is not to say that a (possibly unknown)
+verification object may exist outside the subject's experience (construction) of
+it. Rather, this is a trivial equivalence, since to say that an object exists it
+the same as to say that the subject has constructed it.\footnote{The idea that
+verification or proof objects exist separately from our experience of them is
+part of the realist ontology which is now espoused by Martin-L\"of
+\cite{prawitz:2012}, contrary to his position at the time of the Siena lectures;
+this view of course cannot be accepted by Brouwerians, who profess a thoroughly
+idealist ontology \cite{sundholm-mva, sundholm:2014}.}
 
@@
     \Eval{N}{N'} &
-    \VAL{A'}(\InputMode{M}, \InputMode{N})
+    \VAL{A'}(\InputMode{M'}, \InputMode{N'})
   }\qquad

### 2015-08-01 paper.tex
fill in a missing step
discuss functionality for quantifiers in the theory of sets

@@
 \newcommand\Conj[2]{#1\land #2}
+\newcommand\Disj[2]{#1\lor #2}
+\newcommand\Forall[3]{(\forall #2 \in #1)#3}
+\newcommand\Exists[3]{(\exists #2 \in #1)#3}
 \newcommand\It{\star}
@@
 \newcommand\Pair[2]{\langle #1, #2 \rangle}
-\newcommand\Fst{\mathsf{fst}}
-\newcommand\Ap{\mathsf{ap}}
+\newcommand\Inl[1]{\mathsf{inl}(#1)}
+\newcommand\Inr[1]{\mathsf{inr}(#1)}
+
+\newcommand\BISH{\textbf{BISH}}
 
@@
     \}\\
+  \VAL{\Disj{P}{Q}} &\equiv
+    \{ \InputMode{\Inl{M}}
+    \mid \Member{M}{P}
+    \} \cup
+    \{ \InputMode{\Inr{M}}
+    \mid \Member{M}{Q}
+    \}\\
   \EXP{P} &\equiv
@@
   \GenJ{x}{
+    \HypJ{\Member{\Pair{\It}{\It}}}{\EXP{\False}(\InputMode{x}}
+  }\\
+  \GenJ{x}{
     \HypJ{\Member{\Pair{\It}{\It}}{\True}}{\Eval{x}{M}, \VAL{\False}(\InputMode{M})}
@@
 
-The unary logical relations exactly express the content of a realizability
+The unary logical relations express exactly the content of a realizability
 model, but they still do not yield a theory of sets which is sufficient for
@@
 
+In order to make an important point about functionality, we will define
+intuitionistic existential and universal quantification rather than their
+special cases, conjunction and implication.
+%
 \begin{align*}
@@
   \VAL{\False} &\equiv \{\}\\
-  \VAL{\Imp{P}{Q}} &\equiv 
+  \VAL{\Forall{A}{x}{B}} &\equiv 
     \{ (\InputMode{\Lam{x}{E}}, \InputMode{\Lam{x}{E'}})
     \mid \GenJ{y,z}{
-      \HypJ{\EqMember{[y/x]E}{[z/x]E'}{Q}}{\EqMember{x}{y}{P}}
+      \HypJ{\EqMember{[y/x]E}{[z/x]E'}{[y/x]B}}{\EqMember{x}{y}{A}}
     }
     \}\\
-  \VAL{\Conj{P}{Q}} &\equiv
+  \VAL{\Exists{A}{x}{B}} &\equiv
     \{ (\InputMode{\Pair{M}{N}}, \InputMode{\Pair{M'}{N'}})
-    \mid \EqMember{M}{M'}{P}, \EqMember{N}{N'}{Q}
+    \mid \EqMember{M}{M'}{A}, \EqMember{N}{N'}{[M/x]B}
     \}\\
-  \EXP{P} &\equiv
+  \VAL{\Disj{A}{B}} &\equiv
+    \{ \InputMode{\Inl{M}}
+    \mid \Member{M}{A}
+    \} \cup
+    \{ \InputMode{\Inr{M}}
+    \mid \Member{M}{B}
+    \}\\
+  \EXP{A} &\equiv
     \{ (\InputMode{M}, \InputMode{N})
-    \mid \Eval{M}{M'}, \Eval{N}{N'}, \VAL{P}(\InputMode{M'}, \InputMode{N'})
+    \mid \Eval{M}{M'}, \Eval{N}{N'}, \VAL{A}(\InputMode{M'}, \InputMode{N'})
     \}
 \end{align*}
+
+Now, in contrast to the treatment of the quantifiers in formal intuitionistic
+logic (or in the Mitchell-B\'enabou language of a topos
+\cite{maclane-moerdijk}), it is part of their meaning that their verifications
+should respect the equality of the domain of discourse; this constraint is
+called functionality, and reflects the fact that universal quantification is
+taken as a more general form of implication. Within the theory of sets, there is
+simply not a quantifier which expresses non-functional generality; in this way,
+contrary to the state of affairs in \BISH{} \cite{bishop:1967}, the theorem of
+choice is in fact verified by a choice \emph{function}, not merely a choice
+\emph{operation}.
 
@@
 
-The judgments and demonstration acts are the activity of the subject in
-performing and experiencing mathematics. On the other hand, the theory of sets
-that we have defined is the correct level at which to \emph{do} mathematics,
+The judgments and demonstrations are the activity of the subject in performing
+and experiencing mathematics. On the other hand, the theory of sets that we have
+defined is, to my mind, the correct level at which to \emph{do} mathematics,
 where objects are concrete and have an extensional identity. By guaranteeing
 pervasive functionality, we have embedded in the rich world of intuitionistic
-mathematics a haven safe from the paradoxes that arise from failures of
+mathematics a haven invulnerable to the paradoxes that arise from failures of
 extensionality, such as Diaconescu's theorem \cite{sterling:diaconescu}.
@@
 
-\newpage
-
 \nocite{*}

### 2015-08-01 paper.tex
cite nice van dalen paper

@@
 thence propositional implication? The answer to this question is not clearly
-determinate from the writings of Brouwer and Heyting, since in their lifetimes
-they propounded multiple (sometimes conflicting) explanations of the
-hypothetical judgment. To my mind, the determination of an acceptable
-explanation must be, then, an act of abduction and reconstruction, taking
-into account its adequacy for the expression of the bar theorem and, more
+determinate from the writings of Brouwer, Heyting and Kolmogorov, since in their
+lifetimes they propounded multiple (sometimes conflicting) explanations of the
+hypothetical judgment \cite{van-dalen:2004}. To my mind, the determination of an
+acceptable explanation must be, then, an act of abduction and reconstruction,
+taking into account its adequacy for the expression of the bar theorem and, more
 generally, the development of an open-ended framework for transcendental

### 2015-08-01 paper.tex
wording & typo

@@
 from the sequent judgments $\Seq{\Gamma}{\cdots}$, which are not even
-higher-order judgments at all.} was explained by Martin-L\"of in terms of his
-definition of a hypothetical \emph{proof} or \emph{demonstration}, which is as
-follows:
+higher-order judgments at all.} was explained by Martin-L\"of in terms of
+hypothetical \emph{proof} or \emph{demonstration}, which he defined as follows:
 
@@
   \VAL{\Disj{A}{B}} &\equiv
-    \{ \InputMode{\Inl{M}}
-    \mid \Member{M}{A}
+    \{ (\InputMode{\Inl{M}}, \InputMode{\Inl{N}})
+    \mid \EqMember{M}{N}{A}
     \} \cup
-    \{ \InputMode{\Inr{M}}
-    \mid \Member{M}{B}
+    \{ (\InputMode{\Inr{M}}, \InputMode{\Inr{N}})
+    \mid \EqMember{M}{N}{B}
     \}\\

### 2015-08-01 paper.tex
cite pml's axiom of choice paper

@@
 mathematics a haven invulnerable to the paradoxes that arise from failures of
-extensionality, such as Diaconescu's theorem \cite{sterling:diaconescu}.
+extensionality, such as Diaconescu's theorem \cite{martin-lof:2009, sterling:diaconescu}.
 

### 2015-08-01 paper.tex
...

@@
 $\IsTrue{\Imp{P}{Q}}$ may be asserted if and only if we
-possess a construction $r$, which, joined to any demonstration of $\IsTrue{P}$,
+possess a construction $\mathfrak{r}$, which, joined to any demonstration of $\IsTrue{P}$,
 would automatically effect a demonstration of $\IsTrue{Q}$.

### 2015-08-01 paper.tex
wording

@@
 verification object may exist outside the subject's experience (construction) of
-it. Rather, this is a trivial equivalence, since to say that an object exists it
+it. Rather, this is a trivial equivalence, since to say that an object exists is
 the same as to say that the subject has constructed it.\footnote{The idea that
@@
 A realizability model in this simple sense amounts to interpreting the
-propositions into unary logical relations. Defining a proposition, then, amounts
-to defining the unary relation $\VAL{P}$ (which is the set of verification
+propositions into unary logical relations. To define a proposition, then, is to
+define the unary relation $\VAL{P}$ (which is the species of verification
 objects of the proposition $P$); then, a separate logical relation $\EXP{P}$ is

### 2015-08-01 paper.tex
typo
wordsmithing

@@
 proposition on the other hand? A judgment is an act or an experience, whereas a
-proposition is a mathematical object which may be experienced in different ways:
-for instance, the assertion of the truth of a proposition (i.e.\ $\IsTrue{P}$)
-consists in the fulfillment of the intention expressed by the proposition, and
+proposition is a mathematical object which may be experienced in different ways.
+For instance, the assertion of the truth of a proposition (i.e.\ $\IsTrue{P}$)
+consists in the fulfillment of the intention expressed by the proposition, while
 the recognition of an object as a proposition (i.e.\ $\IsProp{P}$) is the act of
@@
 says ``assert a proposition'', in light of Martin-L\"of's clarification, we must
-understand ``assert \emph{the truth of} a proposition''. Heyting's explanation
+read ``assert \emph{the truth of} a proposition''. Heyting's explanation
 of the assertion conditions for the truth of implication were as follows:
@@
 
-We cannot actually rewrite our definition directly in that style because it is
-impredicative as a definition (as opposed to as a mathematical statement);
-following Dummett \cite{Dummett:Elements} and Martin-L\"of, we must start from a
-distinction between the direct experience of truth (which we shall call
-\emph{verification}) and the indirect experience (which we shall call merely
-\emph{truth}). Then, the intention of a proposition is its verification, and the
-truth of a proposition is the experience of a means of fulfillment for that
-intention:
+It is tempting to rewrite the definition in the following way:
+\begin{quote}
+  (*) To know $\IsTrue{\Imp{P}{Q}}$ is to know $\HypJ{\IsTrue{Q}}{\IsTrue{P}}$.
+\end{quote}
+
+However, as a definition, this is impredicative. Following Dummett
+\cite{Dummett:Elements} and Martin-L\"of, we must start from a distinction
+between the direct experience of truth (which we shall call \emph{verification})
+and the indirect experience (which we shall just call \emph{truth}). Then, the
+intention of a proposition is its verification, and the truth of a proposition
+is the experience of a means of fulfillment for that intention:
 
@@
 which, when adjoined with a construction of the premise, effects a construction
-of the conclusion; and moreover, $\mathfrak{r}$ may proceed by virtue of any of
-the transcendental observations which are possible by virtue of the truth of $P$
-having been experienced by the subject.
+of the conclusion; and moreover, $\mathfrak{r}$ may proceed by appealing to any
+of the transcendental observations which are possible by virtue of the truth of
+$P$ having been experienced by the subject.
 
@@
 
-Why? This purports to be a proof of the judgment
+Why not? This purports to be a proof of the judgment
 $\HypJ{\IsTrue{P}}{\IsTrue{P\land Q}}$, which is not evident under the uniform
@@
 Indeed, Martin-L\"of justifies the elimination rules using the non-uniform
-(material, rather than logical) consequence. This rule, for instance, is
+(material, rather than logical) consequence. The above rule, for instance, is
 justified as follows:
@@
 judgment whose evidence consists in any effective means at all of transforming
-the demonstration of the premise into a demonstration of the conclusion, so the
-use of the uniform hypothetical judgment elsewhere has not somehow saved us the
-need to explain this kind of inference.
+the demonstration of the premise into a demonstration of the conclusion. So, the
+use of the uniform hypothetical judgment elsewhere has not relieved us from the
+need to explain the inference from premise to conclusion in an elimination rule.
 
@@
 
-A construction for a judgment is simply the act of coming to know it---if it is
-to be thought of as an object, it is at least a tensed, ephemeral one; this is
-what Martin-L\"of calls a \emph{demonstration}. On the other hand, a
+A construction for a judgment is simply the act of coming to know it: this is
+what Martin-L\"of calls a \emph{demonstration}, and if it is to be thought of as
+an object, it is at least a tensed, ephemeral one. On the other hand, a
 construction for a proposition is a mathematical object, not an experience: it
@@
 in terms of their verification acts with new explanations in terms of
-verification objects (i.e. their canonical witnesses); then, the verification
+verification objects (i.e.\ their canonical witnesses); then, the verification
 act consists in constructing a verification object.
@@
   \GenJ{x}{
-    \HypJ{\Member{\Pair{\It}{\It}}}{\EXP{\False}(\InputMode{x}}
+    \HypJ{\Member{\Pair{\It}{\It}}{\True}}{\EXP{\False}(\InputMode{x})}
   }\\
@@
 would not have been valid, unless we were to eschew the verificationist meaning
-explanation and also add ``use'' rules (i.e. direct eliminations) for each
+explanation and also add ``use'' rules (i.e.\ direct eliminations) for each
 proposition in addition to the verification rules.
@@
 reasoning about mathematical objects, which have extensional identity. In order
-to consider the equality of of witnesses, the unary logical relations are
-replaced with binary ones (i.e. partial equivalence relations), as follows:
+to consider the equality of of sets and witnesses, the unary logical relations
+are replaced with binary ones (sc.\ partial equivalence relations), as follows:
 

### 2015-08-01 paper.tex
remove citation from abstract

@@
 lifetimes they propounded multiple (sometimes conflicting) explanations of the
-hypothetical judgment \cite{van-dalen:2004}. To my mind, the determination of an
+hypothetical judgment. To my mind, the determination of an
 acceptable explanation must be, then, an act of abduction and reconstruction,
@@
 logic (or in the Mitchell-B\'enabou language of a topos
-\cite{maclane-moerdijk}), it is part of their meaning that their verifications
-should respect the equality of the domain of discourse; this constraint is
-called functionality, and reflects the fact that universal quantification is
-taken as a more general form of implication. Within the theory of sets, there is
-simply not a quantifier which expresses non-functional generality; in this way,
-contrary to the state of affairs in \BISH{} \cite{bishop:1967}, the theorem of
-choice is in fact verified by a choice \emph{function}, not merely a choice
-\emph{operation}.
+\cite{maclane-moerdijk}), in this setting it is part of their meaning that their
+verifications should respect the equality of the domain of discourse; this
+constraint is called functionality, and reflects the fact that universal
+quantification is reconstructed as a more general form of implication. Within
+the theory of sets, there is simply not a quantifier which expresses
+non-functional generality; in this way, contrary to the state of affairs in
+\BISH{} \cite{bishop:1967}, the theorem of choice is in fact verified by a
+choice \emph{function}, not merely a choice \emph{operation}.
 

### 2015-08-01 paper.tex
...

@@
 
-The uniform explanation of hypothetical judgment also deprives Martin-L\"of's
-framework of a propositional connective as strong as proper intuitionistic
-implication. Contra Martin-L\"of's statement in the Siena lectures, the
-interpretation of hypothetical judgment as material consequence is also crucial
-for the semantics of his type theory, as noted by Dybjer \cite{dybjer:testing}.
+Contra Martin-L\"of's explanation in the Siena lectures, the interpretation of
+hypothetical judgment as material consequence is crucial for the semantics of
+his type theory, as noted by Dybjer \cite{dybjer:testing}.
 

### 2015-08-11 paper.tex
typo & cleanup

@@
 thence propositional implication? The answer to this question is not clearly
-determinate from the writings of Brouwer, Heyting and Kolmogorov, since in their
-lifetimes they propounded multiple (sometimes conflicting) explanations of the
-hypothetical judgment. To my mind, the determination of an
-acceptable explanation must be, then, an act of abduction and reconstruction,
-taking into account its adequacy for the expression of the bar theorem and, more
-generally, the development of an open-ended framework for transcendental
-arguments in mathematics.
+determinate from the writings of Brouwer, Heyting and Kolmogorov, since in
+their lifetimes they propounded multiple (sometimes conflicting) explanations
+of the hypothetical judgment. To my mind, the determination of an acceptable
+explanation must take into account its adequacy for the expression of the bar
+theorem and, more generally, the development of an open-ended framework for
+transcendental arguments in mathematics.
 
@@
 reasoning about mathematical objects, which have extensional identity. In order
-to consider the equality of of sets and witnesses, the unary logical relations
+to consider the equality of sets and witnesses, the unary logical relations
 are replaced with binary ones (sc.\ partial equivalence relations), as follows:
@@
     \mid \GenJ{y,z}{
-      \HypJ{\EqMember{[y/x]E}{[z/x]E'}{[y/x]B}}{\EqMember{x}{y}{A}}
+      \HypJ{\EqMember{[y/x]E}{[z/x]E'}{[y/x]B}}{\EqMember{y}{z}{A}}
     }

### 2015-09-01 paper.tex
...

@@
 demonstrations of the antecedent to demonstrations of the consequent) formed the
-backbone of Zeilberger, Harper and Licata's work on focused calculi, which
+backbone of Zeilberger, Harper and Licata's work on higher-order focused calculi, which
 provide a convincing abstract notation for the traces of verification and use

### 2015-10-16 paper.tex
Add section on inference vs consequence

@@
 
+\subsection{What is the difference between an inference rule and a hypothetical judgment?}
+
+Is a rule of inference really the same as a consequence or hypothetical
+judgment, as Martin-L\"of claimed? One way to elucidate the differences
+is to consider them in the context of a Beth or Kripke semantics.
+
+The validity of an inference rule $\infer{\JJ_2}{\JJ_1}$ at a world lies in an
+effective transformation of experiences of $\JJ_1$ at that world to experiences
+of $\JJ_2$ at that world; on the other hand, to experience a hypothetical
+judgment $\HypJ{\JJ_2}{\JJ_1}$ at a world is to have a means to transform
+experiences of $\JJ_1$ at any future world into experiences of $\JJ_2$.
+
+Construed in this way, judgments can be explained by specifying when/where they
+are forced; it is a reasonable requirement that if we shall consider $\JJ$ to
+be a judgment, then for any worlds $u\preceq v$, from $u\Vdash\JJ$ we may
+conclude $v\Vdash\JJ$ (this is called \emph{monotonicity}). The hypothetical
+judgment, at least as explained above, preserves the monotonicity of knowledge
+by definition, whereas rules of inference may not in general satisfy this property.
+
+On the contrary, an "admissible rule" $\infer{\JJ_2}{\JJ_1}$ is sensitive to
+changes in the state of knowledge, and may cease to be valid if a previously
+unknown way to experience $\JJ_1$ is found. Such rules may only be construed as
+hypothetical judgments if the acts specified by the meanings of their premises
+are sufficiently circumscribed so as to satisfy the monotonicity requirement.
+
+Martin-L\"of's identification of the rules of logic with hypothetical judgments, then,
+only obtains because the concepts of \emph{verification} of a proposition, and
+(secondarily) \emph{truth} of a proposition are fixed in advance for all time
+by means of \emph{canonical forms} and \emph{computation} respectively.
+
 \section{Realizability and Type Theory}

### 2018-01-04 paper.tex
...

@@
-\documentclass[11pt]{amsart}
-\usepackage{setspace,url,hyperref}
-\usepackage{amssymb,stmaryrd}
+\documentclass[11pt]{article}
+\usepackage[tt=false]{libertine}
+\usepackage{url,hyperref}
+\usepackage{amsmath,amsthm,amssymb,stmaryrd}
 \usepackage{proof}
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\usepackage{sectsty}
+\allsectionsfont{\sffamily}
+
 \newtheorem{thm}{Theorem}[section]
@@
 \newtheorem{remark}[thm]{Remark}
-\numberwithin{equation}{section}
 
-\def\InputModeColorName{MidnightBlue}
-\def\OutputModeColorName{Maroon}
-\newcommand\InputMode[1]{{\color{\InputModeColorName}{#1}}}
-\newcommand\OutputMode[1]{{\color{\OutputModeColorName}{#1}}}
+\newcommand\InputMode[1]{#1}
+\newcommand\OutputMode[1]{#1}
 
@@
 \author{Jonathan Sterling}
-\address{}
+\date{Fall 2015}
 
-\onehalfspacing
+\maketitle
 
@@
   determinate from the writings of Brouwer, Heyting and Kolmogorov, since in
-their lifetimes they propounded multiple (sometimes conflicting) explanations
-of the hypothetical judgment. To my mind, the determination of an acceptable
-explanation must take into account its adequacy for the expression of the bar
-theorem and, more generally, the development of an open-ended framework for
+  their lifetimes they proposed multiple (sometimes conflicting) explanations
+  of the hypothetical judgment. The determination of an acceptable explanation
+  must take into account its adequacy for the expression of the bar theorem
+  and, more generally, the development of an open-ended framework for
   transcendental arguments in mathematics.
@@
 \end{abstract}
-
-\maketitle
 
@@
 and experiencing mathematics. On the other hand, the theory of sets that we have
-defined is, to my mind, the correct level at which to \emph{do} mathematics,
+defined is the correct level at which to \emph{do} mathematics,
 where objects are concrete and have an extensional identity. By guaranteeing

### 2018-01-04 paper.tex
...

@@
 
-On the contrary, an "admissible rule" $\infer{\JJ_2}{\JJ_1}$ is sensitive to
-changes in the state of knowledge, and may cease to be valid if a previously
-unknown way to experience $\JJ_1$ is found. Such rules may only be construed as
+On the contrary, a rule $\infer{\JJ_2}{\JJ_1}$ merely expresses the closure of
+all mathematical constructions under the inference of $\JJ_2$ from $\JJ_1$;
+this kind of closure is sensitive to changes in the state of knowledge, and may
+cease to be valid if a previously unknown way to experience $\JJ_1$ is found.
+This is the sense in which extra-logical principles like Brouwer's Bar Thesis
+and Markov's Principle are ``rules''~\cite{sundholm-mva}. Such rules may only be construed as
 hypothetical judgments if the acts specified by the meanings of their premises

### 2018-01-04 paper.tex
...

@@
 
-Now, in contrast to the treatment of the quantifiers in formal intuitionistic
-logic (or in the Mitchell-B\'enabou language of a topos
-\cite{maclane-moerdijk}), in this setting it is part of their meaning that their
-verifications should respect the equality of the domain of discourse; this
-constraint is called functionality, and reflects the fact that universal
+Now, in contrast to the treatment of the quantifiers in topos
+logic~\cite{maclane-moerdijk}, in this setting it is part of their meaning that
+their verifications should respect the equality of the domain of discourse;
+this constraint is called functionality, and reflects the fact that universal
 quantification is reconstructed as a more general form of implication. Within
